---
title: "Getting Started and R Nuts and Bolts"
author: "Nathan Young"
date: "`r format(Sys.time(), '%B %d, %Y')`"
output:
  html_document:
    toc: yes
    toc_float:
      toc_collapsed: yes
    theme: spacelab
  pdf_document:
    toc: yes
---

## Vectors and Lists
#### Creating Vectors
```{r Creating Vectors}
x <- c(0.5, 0.6) ## numeric
x <- c(TRUE, FALSE) ## logical
x <- c(T, F) ## logical
x <- c("a", "b", "c") ## character
x <- 9:29 ## integer
x <- c(1 + 0i, 2 + 4i) ## complex
```

Using the `vector()` function:
```{r}
x <- vector("numeric", length = 10)
x
```

#### Mixing Objects
```{r Mixing Objects}
y <- c(1.7, "a") ## character
y <- c(TRUE, 2) ## numeric
y <- c("a", TRUE) ## character
```

#### Explicit Coercion
Objects can be explicity coerced from one class to another using the `as.*` functions, if available.
```{r Explicit Coercion}
x <- 0:6
class(x)
as.numeric(x)
as.logical(x)
as.character(x)
```

Nonsensical coercion results in NAs.
```{r}
x <- c("a", "b", "c")
as.numeric(x)
as.logical(x)
as.complex(x)
```

#### Lists
Lists are a special type of vector that can contain elements of different classes.
```{r Lists}
x <- list(1, "a", TRUE, 1 + 4i)
x
```

## Matrices
Vectors with a *dimension* attribute.

Matrices are constructed *column-wise*, so entries can be thought of starting in the top left corner and running down columns.
```{r Matrices}
m <- matrix(nrow = 2, ncol = 3)
m
dim(m)
attributes(m)
m <- matrix(1:6, nrow = 2, ncol = 3)
m
m <- 1:10
m
dim(m) <- c(2, 5)
m
```

Matrices can be created by column-binding (`cbind()`) or row-binding (`rbind()`).
```{r cbinding and rbinding}
x <- 1:3
y <- 10:12
cbind(x, y)
rbind(x, y)
```

## Factors
Used to represent categorical data.  Can by unordered or ordered.  An integer vector where each integer has a label.
```{r factors}
x <- factor(c("yes", "yes", "no", "yes", "no"))
x
table(x)
unclass(x)
x <- factor(c("yes", "yes", "no", "yes", "no"),
            levels = c("yes", "no"))
x
```

## Missing Values
Missing values are denoted by `NA` or `NaN` for undefined math operations.
```{r missing values}
x <- c(1, 2, NA, 10, 3)
is.na(x)
is.nan(x)
x <- c(1, 2, NaN, NA, 4)
is.na(x)
is.nan(x)
```

## Data Frames
Used to store tabular data.
```{r data frames}
x <- data.frame(foo = 1:4, bar = c(T, T, F, F))
x
nrow(x)
ncol(x)
```

## Names Attribute
Objects can have names.
```{r names object}
x <- 1:3
names(x)
names(x) <- c("foo", "bar", "norf")
x
names(x)
```

Lists can also have names.
```{r names list}
x <- list(a = 1, b = 2, c = 3)
x
```

Matrix names are called `dim names`.
```{r names matrix}
m <- matrix(1:4, nrow = 2, ncol = 2)
dimnames(m) <- list(c("a", "b"), c("c", "d"))
m
```

## Reading Data
#### read.table
One of the most commonly used functions for reading data is [`read.table`](https://www.rdocumentation.org/packages/utils/versions/3.6.2/topics/read.table).

#### Reading Large Data
To make reading large data with `read.table` easier:

- Read the docs
- If data set is larger than RAM, stop.
- Set `comment.char = ""` if there are no commented lines.
- Use `colClasses` command instead of letting R figure data type of column.
- Set `nrows`

## Textual Formats

- `dumping` and `dputing` result in editable text, potentially allowing it to be recovered.
- `dump` and `dput` preserve metadata
- Textual formats work better with version control
- Can be easier to identify issues
- Format is not space efficient

Dput
```{r dputting R objects}
y <- data.frame(a = 1, b = "a")
dput(y)
dput(y, file = "y.R")
new.y <- dget("y.R")
new.y
```

Dump
```{r dumping R objects}
x <- "foo"
y <- data.frame(a = 1, b = "a")
dump(c("x", "y"), file = "data.R")
rm(x, y)
source("data.R")
y
x
```

## Connections
```{r connections, eval=FALSE}
con <- file("foo.txt", "r")
data <- read.csv(con)
close(con)
```

is the same as
```{r eval=FALSE}
data <- read.csv("foo.txt")
```


Reading Lines of a Text File
```{r eval=FALSE}
con <- gzfile("words.gz")
x <- readLines(con, 10)
```

```{r}
con <- url("http://www.jhsph.edu", "r")
x <- readLines(con)
head(x)
```

## Subsetting

- `[` always returns an object of the same class as the original
- `[[` is used to extract elements of a list of data frame
- `$` is used to extract elements of a list or dat frame by name

```{r subsetting}
x <- c("a", "b", "c", "c", "d", "a")
x[1]
x[2]
x[1:4]
x[x > "a"]
u <- x > "a"
u
x[u]
```

#### Subsetting Lists
```{r subsetting lists}
x <- list(foo = 1:4, bar = 0.6)
x[1]
x[[1]]
x$bar
x[["bar"]]
x["bar"]
x <- list(fo = 1:4, bar = 0.6, baz = "hello")
x[c(1, 3)]
name <- "foo"
x[[name]] ## computed index for "foo"
x$name ## element "name" doesn't exist!
x$foo ## element "foo" does exist
```

The `[[` can take an integer sequence.
```{r subsetting nested elements}
x <- list(a = list(10, 12, 14), b = c(3.14, 2.81))
x[[c(1, 3)]]
x[[1]][[3]]
x[[c(2, 1)]]
```

#### Subsetting a Matrix
Matrices can be subsetted in the usual way with `(i, j)` type indices.

Indices can also be missing.
```{r subsetting matrix}
x <- matrix(1:6, 2, 3)
x[1, 2]
x[2, 1]
x[1, ]
x[, 2]
x[1, 2, drop = FALSE]
x[1, , drop = FALSE]
```

#### Partial Matching
Partial matching of names is allowed with `[[` and `$`.
```{r partial matching}
x <- list(aardvark = 1:5)
x$a
x[["a"]]
x[["a", exact = FALSE]]
```

#### Removing NA Values
```{r removing NA values}
x <- c(1, 2, NA, 4, 5)
bad <- is.na(x)
x[!bad]
x <- c(1, 2, NA, 4, NA, 5)
y <- c("a", "b", NA, "d", NA, "f")
good <- complete.cases(x, y)
good
x[good]
y[good]
airquality[1:6,]
good <- complete.cases(airquality)
airquality[good, ][1:6, ]
```

## Vectorized Operations
Many operations in R are *vectorized* making code more efficient, concise, and easier to read.
```{r vectorized ops}
x <- 1:4; y <- 6:9
x + y
x > 2
x >= 2
y == 8
x * y
x / y
x <- matrix(1:4, 2, 2); y <- matrix(rep(10, 4), 2, 2)
x * y ## element-wise multiplication
x / y
x %*% y ## true matrix multiplication
```

















